---
title: "Voting drop boxes"
output: 
  flexdashboard::flex_dashboard:
    orientation: columns
    vertical_layout: fill
---

```{r setup, include=FALSE}
library(sf)
#library(tmap)
library(tidyverse)
#library(leaflet)
#library(ggmap)  # requires API for google map
library(tidygeocoder)
library(here)
library(mapbaltimore)   # map baltimore!
theme_set(theme_void())   # for mapbaltimore
library(getdata)

map_area <- function(x, col) {
  ggplot(data = x) +
    geom_sf(aes(fill = .data[[col]])) +
    geom_sf_label(aes(label = .data[[col]])) +
    guides(fill = "none")
}

library(DT)
```

```{r preprocess data, include=FALSE}
# Load drop boxes address information
here()
drop_boxes <- read_csv(here("Ballot_drop_box_locations_Baltimore_2024.csv"), locale=locale(encoding="latin1"), col_names = FALSE)

# Remove weird character
drop_boxes <- drop_boxes %>%
  mutate(X1 = str_replace_all(X1, "Ê",""))

# Separate into names and street addresses
names_db <- drop_boxes[seq(1,nrow(drop_boxes),2),]
addr_db <- drop_boxes[seq(2,nrow(drop_boxes),2),]

db_df <- data.frame(cbind(names_db, addr_db))
colnames(db_df) <- c("name","addr")

# Convert addresses to lat/lon coordinates (this takes a minutes!)
lat_longs <- db_df %>%
  geocode(addr, method = 'osm', lat = latitude , long = longitude)

# Check for mistakes and fix them
# missing
nan_inds <- which(is.na(lat_longs$latitude))
lat_longs[nan_inds,]

# Carver Vocational Tech
lat_longs$latitude[nan_inds[1]] <- 39.30446
lat_longs$longitude[nan_inds[1]] <- -76.65264

# Farring Baybrook Rec Center
lat_longs$latitude[nan_inds[2]] <- 39.22514
lat_longs$longitude[nan_inds[2]] <- -76.59747

# out of bounds
oob <- which(lat_longs$latitude > 39.5 | lat_longs$latitude < 39)
lat_longs[oob,]

# Baltimore Museum of Art
lat_longs$latitude[oob[1]] <- 39.32614
lat_longs$longitude[oob[1]] <- -76.61950

# League for people with disabilities
lat_longs$latitude[oob[2]] <- 39.34480
lat_longs$longitude[oob[2]] <- -76.59856

# Check ranges now
range(lat_longs$latitude)
range(lat_longs$longitude)

#saveRDS(lat_longs, "drop_box_lat_lon.RDS")

```

```{r load data}
# Load clean data
lat_longs <- readRDS("drop_box_lat_lon.RDS")
```
## Where can I drop by ballot?

Election season is upon us already, and I was super pumped to participate in the new vote by mail program in Baltimore! As you probably already know, you can return your ballot by mail (just drop it in any USPS box), or if you'd like to save the City a little money, you can return your ballot to any of 33 drop boxes around Baltimore City. But where exactly are these boxes, you ask? The addresses are posted online, but as a newbie to this town, it wasn't immediately obvious which one was the closest to me...so I made a map. I hope it helps make your voting experience a little easier too.

```{r build map}
# Load data by "neighborhood" (or council district)
area <- get_baltimore_area(
  type = "neighborhood"
)

layer_streets <- maplayer::layer_location_data(
  location = area,
  data = streets,
  color = "gray60"
)

layer_parks <- maplayer::layer_location_data(
  location = area,
  data = parks,
  fill = "forestgreen",
)

background_layers <- list(layer_streets, layer_parks)


```

```{r}
# convert coordinates to sf object
 DT_sf = st_as_sf(lat_longs, coords = c("longitude", "latitude"), 
                  crs = 4326, agr = "constant")

# Add "number" column to use as label below
DT_sf2 <- DT_sf %>%
  mutate(number = 1:nrow(DT_sf))

bmore_map +
  layer_markers(lat_longs, color="red") +
  layer_labelled(
    data = DT_sf2,
    geom = "sf_label",
    mapping = aes(label = number),
  )


```

```{r}
# Make pretty table
lat_longs_pretty <- lat_longs %>%
  select(Name = name,
         Address = addr) 
datatable(lat_longs_pretty)

```
